---
title: "Homework: Simulation"
author: "Firstname Lastname"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3)
```

## Instructions

When you are finished with the assignment:

1. Save the file as 09_Lastname_Firstname.Rmd and compile it
2. Commit the Rmd file and corresponding html file to your homework git repo
3. Push the commit to github
4. Locate the Rmd file on github and click on the SHA link for the file
5. Paste the url of the resulting page into Canvas to submit your assignment.

Your assignment must compile on a different computer as long as the saspath is set correctly for that machine.
This means you will need to use a local file path when you read the data in via R (and SAS).

## Is it in the cards?

In this task, you will create a simulation of a card or dice game of your choice. You might consider games such as [set](https://en.wikipedia.org/wiki/Set_(card_game)), [farkle](https://www.dicegamedepot.com/farkle-rules/), [blackjack](https://bicyclecards.com/how-to-play/blackjack/#filter), [war](https://bicyclecards.com/how-to-play/war/#filter), or [go fish](https://bicyclecards.com/how-to-play/go-fish/#filter). You may choose any game that can be assessed probabilistically -- you may want to avoid games which require a large amount of strategy, skill, and bluffing, such as poker. 

You don't need to necessarily simulate the full game mechanism, but you do need to implement enough of the game mechanism that you can identify an interesting probability question in your game and answer the question using a simulation.

### Scenario Description

When I was a kid, my parents taught my siblings and I to play hearts by describing the rules: the goal was to get rid of all of your hearts by burying them in other tricks.  

Bicycle Cards describes the rules as follows: 

> OBJECT OF THE GAME
To be the player with the lowest score at the end of the game. When one player hits the agreed-upon score or higher, the game ends; and the player with the lowest score wins.

> CARD VALUES/SCORING
> At the end of each hand, players count the number of hearts they have taken as well as the queen of spades, if applicable. Hearts count as one point each and the queen counts 13 points.    
Each heart - 1 point    
The Q - 13 points    
The aggregate total of all scores for each hand must be a multiple of 26.    
The game is usually played to 100 points (some play to 50).    
When a player takes all 13 hearts and the queen of spades in one hand, instead of losing 26 points, that player scores zero and each of his opponents score an additional 26 points.

At any rate, my parents left off the last part of the scoring mechanism, figuring that it was too complicated strategically for my siblings and I (I was the oldest and was about 11 at the time, so my younger brother would have been 6 or so). 

We started playing, and I ended up crying as we got to the end of the round, because I had collected every possible heart. I was *so mad* that despite my best attempts at strategy, I had lost so terribly.

When they counted the cards up, I had accidentally "shot the moon", and my dad got to explain that I'd actually won that round instead of losing it horribly.

What I'd like to know is whether or not my parents actually made a reasonable call - what is the probability that a completely naive player who doesn't know the full set of rules of the game accidentally shoots the moon? (I'm assuming that at 11 my strategic skills were not developed enough to do more than simple optimization of strategy).

### Set up your objects

> First, you must set up a data frame describing all possible values of your playing items. For instance, if you are using a standard deck of 52 playing cards, you'll need a data frame which has columns for suit and value, and you may want to make value a factor variable (your call on that one). If your game requires that players hold cards in their hand, you may also want to have a column indicating location (deck, discard, player_x). 

```{r}
library(tidyverse)
# make the cards ordered correctly
card_values <- factor(1:13, labels = c(2:10, "Jack", "Queen", "King", "Ace"))
suit <- c("hearts", "diamonds", "spades", "clubs")

deck_of_cards <- tidyr::crossing(suit, value = card_values) %>%
  mutate(points = (suit == "hearts") + 
           13*(suit == "spades" & value == "Queen"))
# it's often easier to add two logical statements 
# together and weight one of them than to create a 
# compound if function or a table that you could 
# use to make another join.
```

I've created a standard deck of cards, and assigned each card the point value it has when scoring a game of hearts. 


### Write one or more functions to simulate the basic game mechanisms

> For instance, if you are simulating set, you need to have a function that draws 12 cards from the deck. You may want an auxiliary function which determines if a set is present.

```{r}
deal_hands <- function(deck, players = 1:4) {
  # Must have between 3 and 5 players"
  stopifnot(length(players) >= 3, length(players) <= 5)
  
  # Rules to ensure that deck is evenly divisible
  # I'm deviating slightly from the official Bicycle rules to make things a bit easier
  if (length(players) == 3) {
    deck <- deck %>%
      filter(!(suit == "diamonds" & value == "2"))
  } else if (length(players) == 5) {
    deck <- deck %>%
      filter(!(suit %in% c("diamonds", "spades") & value == "2"))
  }
  
  # Now, assign players to cards (or cards to players) and sort hands based
  # on value and suit
  deck %>%
    # shuffle the rows
    slice_sample(n = nrow(deck)) %>%
    mutate(player = rep(players, length.out = n())) %>%
    arrange(player, suit, value)
}

set.seed(20201013) # set the seed for reproducibility
test_deck <- deal_hands(deck_of_cards)
```

This function assigns a player to each card in the deck. From here, we could split the data frame up in a number of different ways (e.g. have player-specific data frames) but for now, lets not.


### Identify a probability question you can answer with your simulation

> For instance, if you are simulating set, you may want to know the probability that an array of 12 cards does not contain a set. (You can check your answers by reading the rules of the game, if you want). If you are simulating blackjack, you may want to know what the [expected return on a specific strategy is](https://towardsdatascience.com/python-blackjack-simulator-61b591ffb971) (hint: if you're beating the dealer, your simulation is probably wrong). 

Fundamentally, my question is what the probability is that a player using relatively little strategy accidentally "shoots the moon". 

To do that, I have to write a few more functions to describe legal play:

1. Hearts can't be played as the first card until either:
  - the queen of spades is played, or
  - the player only has hearts left
2. Players must follow suit when possible.

Other than those rules, I am going to avoid adding any other strategic components: players will randomly select from cards which can be legally played to determine what card is played first


### Design a simulation to answer the question you identified

Run your simulation with an increasing number of iterations and create some sort of plot showing your results. 

Describe any issues you encountered in your simulation, and how you addressed them (or where the whole thing went off the rails, if you couldn't address everything in time). 



#### Functions to handle playing hearts

##### Selecting a card to play

I started out with this function that handles the logic of choosing which card to play once the lead suit is known for the trick. 
```{r}
pick_card <- function(cards, sel_suit, hearts_avail = F) {
  if (sel_suit %in% cards$suit) {
    # 2 of clubs is always selected first; if that's an option it is the first trick
    club2 <- tibble(suit = "clubs", value = "2")
    play <- inner_join(cards, club2, by = c("suit", "value"))
    
    # If there is no 2 of clubs, then play according to the suit
    if (nrow(play) == 0) 
    play <- filter(cards, suit == sel_suit) %>% 
      slice_sample(n = 1)
  } else {
    # Does the player have something besides hearts to play?
    non_hearts <- length(setdiff(cards$suit, "hearts")) > 0
    if (!hearts_avail & non_hearts) {
      # if hearts aren't available and there are other suits, draw from those
      play <- filter(cards, suit != "hearts") %>%
        slice_sample(n = 1)
    } else {
      # Otherwise, player can do whatever they want
      play <- slice_sample(cards, n = 1)
    } 
  }
  return(play)
}
```

Then I tested this function out a bit...
```{r}
# player 3 has the 2 of clubs
filter(test_deck, player == 3) %>%
pick_card("clubs")

# What will player 1 play?
# replicate lets us repeat a scenario 50 times to see what happens each time.
# Each result is a tibble, so if we tell it not to simplify the results into a vector,
# we can just bind all of the rows together afterwards.
replicate(50, {
  filter(test_deck, player == 1) %>%
    pick_card(sel_suit = "clubs")
}, simplify = F) %>%
  bind_rows() %>%
  # Now group by basically everything and count up the replications to see
  # if there is anything that's significant
  group_by(suit, value, points, player) %>%
  count()
```

The end result is that player 1 never plays anything but clubs, which is proper given that there are clubs available and they must play in suit. So far, so good.

But, this doesn't help if there is no suit selected (i.e. when we're the lead player and trying to decide what card to play). In the very first trick, clubs is always the lead suit, but after that, we have to be a bit trickier. Rather than writing two functions, can we write a single function that will handle both situations gracefully?

```{r}
pick_card <- function(cards, sel_suit = NA, hearts_avail = F) {
  # add default sel_suit = NA to indicate not selected - this mostly applies
  # to the first card played in each trick that isn't the first trick
  if (is.na(sel_suit)) {
      # if hearts aren't available, filter them out
      if (!hearts_avail) cards <- filter(cards, suit != "hearts")
      
      # then select a card randomly
      play <- slice_sample(cards, n = 1)
    
    return(play)
  }
  
  # If suit is selected we use slightly different logic:
  if (sel_suit %in% cards$suit) {
     # 2 of clubs is always selected first; if that's an option it is the first trick
    club2 <- tibble(suit = "clubs", value = "2")
    play <- inner_join(cards, club2, by = c("suit", "value"))
    
    # If there is no 2 of clubs, then play according to the suit
    if (nrow(play) == 0) {
      play <- filter(cards, suit == sel_suit) %>% 
        slice_sample(n = 1)
    }
  } else {
    # Does the player have something besides hearts to play?
    non_hearts <- length(setdiff(cards$suit, "hearts")) > 0
    if (!hearts_avail & non_hearts) {
      # if hearts aren't available and there are other suits, draw from those
      play <- filter(cards, suit != "hearts") %>%
        slice_sample(n = 1)
    } else {
      # Otherwise, player can do whatever they want
      play <- slice_sample(cards, n = 1)
    } 
  }
  return(play)
}
```

But this function logic is getting complicated and hard to untangle. Let's rewrite it in a more modular way:
```{r}
# Check whether the 2 of clubs is available
pick_first <- function(cards) {
  # Check for 2 of clubs
  # 2 of clubs is always selected first; if that's an option it is the first trick
  inner_join(cards, 
             tibble(suit = "clubs", value = "2"), 
             by = c("suit", "value"))
}

# Check whether there's a card in the selected suit
pick_card_in_suit <- function(cards, sel_suit = NA) {
  # if sel_suit is NA, this will return nothing
  filter(cards, suit == sel_suit) %>%
  slice_sample(n = 1)
}

# Check whether there's a card in another suit
pick_card_out_suit <- function(cards, hearts_avail = F) {
  card_options <- cards
  onlyhearts <- sum(cards$suit != "hearts") > 0
  # If hearts aren't available and there are other suits, filter out the hearts
  if (!onlyhearts | hearts_avail) {
    card_options <- cards %>% filter(suit != "hearts")
  }
  
  slice_sample(card_options, n = 1)
}

pick_card <- function(cards, sel_suit = NA, hearts_avail = F) {
  play <- pick_first(cards)
  
  # Try picking a card of the selected suit
  if (nrow(play) == 0) {
    play <- pick_card_in_suit(cards, sel_suit)
  }
  
  # Otherwise, try picking a card outside of the selected suit
  if (nrow(play) == 0) {
    play <- pick_card_out_suit(cards, hearts_avail)
  }
  
  return(play)
}

```

This is about the same amount of code, but it's much more readable in several smaller functions.

To test this function out it's useful to play out a fragment of a game inside a single code chunk (this will help us write the function, when we need to do so in a bit):

```{r}
# First, lets simulate the first round:
hands <- test_deck %>%
  nest(hand = -player)
# This creates a column named "hand" where each entry is a data frame containing
# the cards in a single player's hand.

round1 <- tibble() # start with an empty tibble
for (df in hands$hand) {
  round1 <- bind_rows(round1, pick_card(df, "clubs", hearts_avail = F))
}
round1 <- round1 %>%
  mutate(player = hands$player)

# Then we remove round1 from the deck:
test_deck_round2 <- anti_join(test_deck, round1)

# And determine who is leading the next round:
lead_round2 <- round1 %>% arrange(desc(value)) %>% slice(1) %>% pluck("player")

lead_card_options <- replicate(50, {
  filter(test_deck, player == lead_round2) %>%
    pick_card(sel_suit = NA)
}, simplify = F) %>% 
  bind_rows() %>%
  group_by(suit, value) %>%
  count()

lead_card_options

# What didn't get selected?
filter(test_deck, player == lead_round2) %>%
  anti_join(lead_card_options)

# As expected, the only card not chosen was the heart, 
# and since players can't lead with hearts yet, 
# everything appears to be working as expected.
```

Now that we've checked that the first card of a trick works, let's go ahead and check to make sure that the remaining cards selected by the other players are reasonable:

```{r}
hands <- test_deck_round2 %>%
  nest(hand = -player)

# Define a quick function to replicate the card sampling process a few times
# and aggregate the results
quick_card_sample_aggregate <- function(handx, suitx, heartsx = F) {
  replicate(50, pick_card(handx, suitx, heartsx), simplify = F) %>%
    bind_rows() %>%
    group_by(suit, value) %>%
    count()
}

# This is the faster way to go - use the function above and then just see
# what the aggregate results look like...
crossing(player = c(1, 3, 4), suit_sel = suit) %>%
  mutate(
    # map2 takes two arguments and passes them to a function
    # in this case, the hand data frame and the suit selection
    card_selected = purrr::map2(hands$hand[player], suit_sel, .f = quick_card_sample_aggregate)
  ) %>%
  unnest(card_selected) %>%
  group_by(suit_sel, suit) %>%
  count(wt = n)
```
This is just testing to see that in round 2, with the initial test deck we have, there aren't any players who can't match the initially played suit. 

My code used a convenience function and the `map2` function from the purrr package. This saves a lot of typing, but it isn't something you've seen formally yet. 
[Here](round-2-test-sampling-straightforward.R) is a way to do it using things that you mostly have seen before, but it's a lot more typing and a lot less organized.

In the first couple of tricks, it is likely that each player has all 4 suits of cards. Let's create a different scenario to ensure that the function performs correctly as we've been using it:

```{r}
last_3_tricks <- test_deck_round2 %>%
  group_by(player) %>%
  sample_n(3) # 3 cards for each player

# This should pick a card from each player because we are passing in a grouped tibble
pick_card(last_3_tricks, sel_suit = "diamonds")
# But it doesn't because not all players have diamonds available 
# (and it just silently skips those, yikes!)

# This picks a card from each player individually
last_3_tricks %>% 
  nest(hand = -player) %>%
  mutate(hand = purrr::map(hand, ~pick_card(., sel_suit = "diamonds", hearts_avail = T))) %>%
  unnest(hand)
```

The pick_card function seems like it works with grouped data (picking one card per player), but that is an illusion, because if one player doesn't have the right suit of cards, it doesn't work properly. Let's write a `pick_indiv_cards` function that will handle things correctly (we'll use the `pick_cards` function inside of this function so we don't have to redo the logic):

```{r}
pick_player_cards <- function(hand, sel_suit = NA, hearts_avail = F) {
  res <- hand %>%
    nest(hand = -player)
  
  # I'm writing this without purrr because you don't know it yet
  for (i in 1:nrow(res)) {
    res$hand[[i]] <- pick_card(res$hand[[i]], sel_suit = sel_suit, hearts_avail = hearts_avail)
  }
  
  res %>%
    unnest(hand)
}

pick_player_cards(last_3_tricks, "diamonds", hearts_avail = T)
```

This behaves the way we expect.

Note, this isn't the most efficient way to write this code. I'm letting it evolve as I write it, but I'm not really factoring in computational efficiency yet. You can worry about efficiency after you have working code and functions that make sense. 

##### Playing a full trick

```{r}
# for this function, we assume deck is going to have a player for each card in the deck
# and that cards which have been played already will have been removed
pick_trick <- function(deck, lead_player = NA, hearts_avail = F) {
  # Check that everyone has a card to play
  stopifnot(nrow(deck) >= length(deck$player))
  
  # lead player is not determined for first trick - it's person with 2 of clubs
  if (is.na(lead_player)) {
    # use pick_card here because we only care about a single player
    lead_card <- pick_card(deck, sel_suit = "clubs")
    lead_player <- lead_card$player
  } else {
    lead_card <- deck %>% filter(lead_player == player) %>% pick_card()
  }
  
  # assess whether hearts can be played based on the lead card
  hearts_avail <- hearts_avail | (lead_card$points > 0)
  
  # collect the remaining cards
  # technically, we should do this player-by-player, but the only difference
  # is whether hearts can be played or not. Lets hand-wave and make use of 
  # the fact that slice_sample() works with groups so that we don't have to
  # specify the order of play...
  trick <- deck %>% 
    filter(player != lead_player) %>% 
    group_by(player) %>% 
    # use pick_indiv_cards here because there are multiple players
    pick_indiv_cards(sel_suit = lead_card$suit, hearts_avail = hearts_avail)
  
  # combine the cards together
  trick <- bind_rows(lead_card, trick) %>%
    mutate(in_suit = (suit == lead_card$suit)) %>%
    # order so that the first card in the trick is the winner of this hand
    arrange(desc(in_suit), desc(value)) %>%
    # keep track of who has which cards
    mutate(cards_to = player[1]) %>%
    # also keep track of whether hearts can be lead yet
    mutate(hearts_avail = hearts_avail | sum(points) > 0)
  
  return(trick)
}
```

Now we have a function to play a trick, so let's see if it works by testing out the first trick:
```{r}
trick_1 <- pick_trick(test_deck, lead_player = NA)
trick_1
```
We can keep going...
```{r}
trick_2 <- test_deck %>% 
  # Remove cards from deck that were played in trick 1
  anti_join(trick_1, by = c("suit", "value")) %>%
  # Get cards for a new trick
  pick_trick(lead_player = trick_1$player[1], hearts_avail = trick_1$hearts_avail[1])
trick_2


trick_3 <- test_deck %>% 
  # Remove cards from deck that were played in trick 1 and 2
  anti_join(trick_1, by = c("suit", "value")) %>%
  anti_join(trick_2, by = c("suit", "value")) %>%
  # Get cards for a new trick
  pick_trick(lead_player = trick_2$player[1], hearts_avail = trick_2$hearts_avail[1])
trick_3

# Might be easier to keep track of cards in a "played" and "unplayed" deck
played <- bind_rows(trick_1, trick_2, trick_3)
unplayed <- anti_join(test_deck, played, by = c("suit", "value"))

trick_4 <- unplayed %>%
  # Get cards for a new trick
  pick_trick(lead_player = trick_3$player[1], hearts_avail = sum(played$hearts_avail) > 0)
trick_4

```

##### Playing a full game

First, we need a function that will keep track of the deck for us - basically, we need a played and unplayed data frame, and then a counter for who is responsible for leading the trick. If at any point the played deck has a point card, then hearts are available for future tricks.

It may also help to keep track of what trick we're on, so let's add the logic for that as well.

```{r}
play_trick <- function(unplayed, played, lead_player) {
  if (nrow(played) == 0) {
    trick_num <- 1
  } else {
    trick_num <- max(played$trick_num) + 1
  }
  
  hearts_avail <- sum(played$points) > 0
  current_trick <- pick_trick(unplayed, 
                              lead_player = lead_player, 
                              hearts_avail = hearts_avail) %>%
    mutate(trick_num = trick_num)
  
  unplayed <- anti_join(unplayed, current_trick, by = c("suit", "value"))
  played <- bind_rows(played, current_trick)
  lead = unique(current_trick$cards_to)
  
  return(list(unplayed = unplayed,
              played = played,
              lead_player = lead))
}


card_values <- factor(1:13, labels = c(2:10, "Jack", "Queen", "King", "Ace"))
suit <- c("hearts", "diamonds", "spades", "clubs")

game_of_hearts <- function(players = 1:4) {
  # start with the full deck
  deck_of_cards <- tidyr::crossing(suit, value = card_values) %>%
    mutate(points = (suit == "hearts") + 
             13*(suit == "spades" & value == "Queen"))
  
  hands <- deal_hands(deck_of_cards, players = players)
  
  # get total number of tricks:
  ntricks <- hands %>% group_by(player) %>% count() %>% pluck("n") %>% unique()
  
  res <- list(unplayed = hands, played = tibble(), lead_player = NA)
  for (trick in 1:2) {
    res <- play_trick(res$unplayed, res$played, res$lead_player)
  }
  
  return(res$played)
}

game_of_hearts()
```

