---
title: "Hearts Simulation"
author: "Firstname Lastname"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3)
```

## Instructions

When you are finished with the assignment:

1. Save the file as 09_Lastname_Firstname.Rmd and compile it
2. Commit the Rmd file and corresponding html file to your homework git repo
3. Push the commit to github
4. Locate the Rmd file on github and click on the SHA link for the file
5. Paste the url of the resulting page into Canvas to submit your assignment.

Your assignment must compile on a different computer as long as the saspath is set correctly for that machine.
This means you will need to use a local file path when you read the data in via R (and SAS).

## Is it in the cards?

In this task, you will create a simulation of a card or dice game of your choice. You might consider games such as [set](https://en.wikipedia.org/wiki/Set_(card_game)), [farkle](https://www.dicegamedepot.com/farkle-rules/), [blackjack](https://bicyclecards.com/how-to-play/blackjack/#filter), [war](https://bicyclecards.com/how-to-play/war/#filter), or [go fish](https://bicyclecards.com/how-to-play/go-fish/#filter). You may choose any game that can be assessed probabilistically -- you may want to avoid games which require a large amount of strategy, skill, and bluffing, such as poker. 

You don't need to necessarily simulate the full game mechanism, but you do need to implement enough of the game mechanism that you can identify an interesting probability question in your game and answer the question using a simulation.

### Scenario Description

When I was a kid, my parents taught my siblings and I to play hearts by describing the rules: the goal was to get rid of all of your hearts by burying them in other tricks.  

Bicycle Cards describes the rules as follows: 

> OBJECT OF THE GAME
To be the player with the lowest score at the end of the game. When one player hits the agreed-upon score or higher, the game ends; and the player with the lowest score wins.

> CARD VALUES/SCORING
> At the end of each hand, players count the number of hearts they have taken as well as the queen of spades, if applicable. Hearts count as one point each and the queen counts 13 points.    
Each heart - 1 point    
The Q - 13 points    
The aggregate total of all scores for each hand must be a multiple of 26.    
The game is usually played to 100 points (some play to 50).    
When a player takes all 13 hearts and the queen of spades in one hand, instead of losing 26 points, that player scores zero and each of his opponents score an additional 26 points.

At any rate, my parents left off the last part of the scoring mechanism, figuring that it was too complicated strategically for my siblings and I (I was the oldest and was about 11 at the time, so my younger brother would have been 6 or so). 

We started playing, and I ended up crying as we got to the end of the round, because I had collected every possible heart. I was *so mad* that despite my best attempts at strategy, I had lost so terribly.

When they counted the cards up, I had accidentally "shot the moon", and my dad got to explain that I'd actually won that round instead of losing it horribly.

What I'd like to know is whether or not my parents actually made a reasonable call - what is the probability that a completely naive player who doesn't know the full set of rules of the game accidentally shoots the moon? (I'm assuming that at 11 my strategic skills were not developed enough to do more than simple optimization of strategy).

### Set up your objects

> First, you must set up a data frame describing all possible values of your playing items. For instance, if you are using a standard deck of 52 playing cards, you'll need a data frame which has columns for suit and value, and you may want to make value a factor variable (your call on that one). If your game requires that players hold cards in their hand, you may also want to have a column indicating location (deck, discard, player_x). 

```{r}
library(tidyverse)
create_deck <- function() {
  # Set up cards
  suits <- c("hearts", "diamonds", "spades", "clubs")
  values <- factor(1:13, labels = c(2:10, "Jack", "Queen", "King", "Ace"))
  deck <- crossing(suit = suits, value = values)

  # Set up points
  deck %>%
    mutate(points = (suit == "hearts") +
             13*(suit == "spades" & value == "Queen")) %>%
    # add a card ID
    mutate(id = 1:n())
}

create_deck()
```

I've created a standard deck of cards, and assigned each card the point value it has when scoring a game of hearts. 

In the Bicycle rules, there are specific cards which are removed if there are 3 or 5 players (to make sure there are an even number of cards for each player). I'm deviating slightly from those rules so that the 2 of clubs is always available (this makes my life easier), but the fundamentals are the same. 

```{r}

deck_fix <- function(deck = create_deck(), n_players = 4) {

  # Remove cards from deck to make the game playable by 3 or 5 players
  if (n_players == 4) {
    # do nothing special
  } else if (n_players == 3) {
    deck <- deck %>%
      filter(!(suit == "diamonds" & value == "2"))
  } else if (n_players == 5) {
    deck <- deck %>%
      filter(!(suit %in% c("diamonds", "spades") & value == "2"))
  } else {
    stop("Hearts is playable by between 3 and 5 players")
  }

  deck
}

create_deck() %>%
  deck_fix(n_players = 3) %>%
  arrange(value, suit)
```




### Write one or more functions to simulate the basic game mechanisms

#### Deal cards to players

```{r}
deal_cards <- function(deck, players = 1:4) {
  deck %>%
    # Fix the deck
    deck_fix(n_players = length(players)) %>%
    # shuffle the rows
    slice_sample(n = nrow(deck)) %>%
    # Assign cards to players one-by-one
    mutate(player = rep(players, length.out = n())) %>%
    # Sort deck by player, suit, and value
    arrange(player, suit, value)
}

create_deck() %>%
  deal_cards(players = c("Amy", "Joe", "Sarah", "Bill")) %>%
  group_by(player) %>%
  count()
```

#### Picking Cards

Next, we need to handle how a player would select a card. We're going to ignore any sort of logic that would depend on cards played by previous players in the same trick (or previous tricks). Remember, I'm simulating my siblings and I playing this game when I was at most 11 (and I was the oldest), so it's a 100% safe assumption that our logical skills were ... suboptimal.

In Hearts, like in many other trick-based games, you have certain rules that must be followed, like matching suits.

```{r}

# Check whether there's a card in the selected suit
pick_card_in_suit <- function(cards, sel_suit = NA) {
  # if sel_suit is NA, this will return nothing
  filter(cards, suit == sel_suit) %>%
    slice_sample(n = 1)
}

```

In addition, hearts requires that if no point cards have been played yet, you must select a card which is not a point card (so something other than a heart or the queen of spades) if you have the option to do so. 

We handle this using the `hearts_available` parameter, which describes whether point cards have been played. If hearts aren't available, and there are other suits in the set of cards, we will pick a non-heart card.

```{r}
# Check whether there's a card in another suit
pick_card_out_suit <- function(cards, hearts_avail = F) {
  if (is.na(hearts_avail)) stop("hearts_avail cannot be NA")
  card_options <- cards

  nonhearts <- any(cards$suit != "hearts")

  # If hearts aren't available and there are other suits,
  # filter out the hearts
  if (nonhearts & !hearts_avail) {
    card_options <- cards %>% filter(suit != "hearts")
  }

  # print(card_options)
  slice_sample(card_options, n = 1)
}

```

(This logic caused 90% of the trouble I had with this game simulation)

Now that we have that, we can write a function that will take a set of cards, and pick a card for a given player in the selected suit (if provided). 

```{r}
pick_card <- function(deck,
                      this_player = NA,
                      sel_suit = NA,
                      hearts_avail = F) {
  if (!is.na(this_player)) {
    options <- filter(deck, player %in% this_player)
  } else {
    options <- deck
  }

  play <- bind_rows(
    # First, try to pick something in suit
    pick_card_in_suit(options, sel_suit = sel_suit),
    # Then if that fails pick something out of suit
    pick_card_out_suit(options, hearts_avail = hearts_avail)
  )

  # Take the first row - in-suit if it's available, out-of-suit otherwise
  play[1,]
}
```


#### Playing the first trick

The first trick is a bit different, logic wise. The person with the 2 of clubs leads (in other tricks, it's the person who took the previous trick). 

To make this function a bit easier to use, I've created a couple of helper functions that can be re-used in any trick:

```{r}
# Evaluate a played trick, adding columns for trick_order, in_suit, player_win, and trick number
eval_trick <- function(cards, spec_suit, trick_num = NA) {
  cards %>%
    mutate(trick_order = 1:n()) %>%
    mutate(in_suit = suit == spec_suit) %>%
    arrange(desc(in_suit), desc(value)) %>%
    mutate(player_win = .$player[1]) %>%
    arrange(trick_order) %>%
    mutate(trick = trick_num)
}

# Remove cards from deck that have been played and bind on as a played trick
fix_deck(dealt, trick) {
  dealt %>%
    anti_join(trick, by = c("suit", "value", "player", "id")) %>%
    bind_rows(trick)
}


```

This leads to the code for the first trick:

```{r}
first_trick <- function(dealt) {
  # Which player has 2 of clubs?
  first_card <- dealt %>%
    filter(suit == "clubs" & value == "2")

  remaining_players <- unique(dealt$player) %>%
    setdiff(first_card$player)

  play <- first_card
  for (i in remaining_players) {
    play <- dealt %>%
      filter(player == i) %>%
      pick_card(sel_suit = "clubs") %>%
      bind_rows(play)
  }
  play <- play %>%
    eval_trick(spec_suit = "clubs",
               trick_num = 1)

  if (!all(dealt$player %in% play$player)) {
    stop("Not all players picked cards?")
  }

  fix_deck(dealt, play)
}
```

#### General trick

The same basic logic applies in any trick - the first card is picked, and then all other players throw in a card. We do need a hearts_playable function that will evaluate our deck to see if hearts is an option; once we have that, we can write a general trick-playing function.

Notice that we're calling the first_trick function within this function - this makes it much easier to remember the syntax. 
```{r}

hearts_playable <- function(df) {
  sum(df$points, na.rm = T) > 0
}

play_trick <- function(dealt) {
  if (!"trick" %in% names(dealt)) {
    return(first_trick(dealt))
  }

  # get empty df
  play <- filter(dealt, id < 0)

  # get trick number
  trick_num <- (1 + dealt$trick) %>% max(na.rm = T)

  # get player to lead
  avail <- filter(dealt, is.na(trick))
  last_player <- filter(dealt, trick == max(trick, na.rm = T)) %>%
    pluck("player_win") %>%
    unique()

  # get already played cards and see if hearts can be played
  already_played <- anti_join(dealt, avail, by = c("suit", "value"))
  can_play_hearts <- bind_rows(already_played, play) %>%
    hearts_playable()

  # pick first card
  play <- pick_card(avail,
                    this_player = last_player,
                    sel_suit = NA,
                    hearts_avail = can_play_hearts)

  # pick remaining cards
  remaining_players <- unique(dealt$player) %>%
    setdiff(last_player)
  for (i in remaining_players) {
    can_play_hearts <- bind_rows(already_played, play) %>%
      hearts_playable()
    play <- bind_rows(
      play,
      pick_card(avail, this_player = i,
                sel_suit = play$suit[1],
                hearts_avail = can_play_hearts)
    )
  }

  # evaluate the trick
  play <- play %>%
    eval_trick(spec_suit = play$suit[1], trick_num)

  # remove trick from deck and add it back in as a played trick
  fix_deck(dealt, play)
}

create_deck() %>%
deal_cards(players = c("Amy", "Joe", "Sarah", "Bill")) %>%
play_trick() %>%
play_trick() %>%
filter(!is.na(trick)) # first 2 tricks in a game
```

#### Play a full game
Now, we can play a full game of hearts by repeatedly calling the play_trick function. All of the cards stay in the same data frame, and we keep track of who won each trick and who played each card - this allows us to calculate the score for each player at the end. 
```{r}
hearts <- function(players) {
  deck <- create_deck() %>%
    deal_cards(players)

  number_of_rounds <- nrow(deck)/length(players)

  for (i in 1:number_of_rounds) {
    # print(i)
    deck <- play_trick(deck)
    # print(filter(deck, trick == i))
  }

  return(deck)
}

hearts(c("Amy", "Beth", "Cara", "Dawn"))
```

#### Calculate game_score
It's pretty easy given our setup to calculate each player's score:
```{r}
calc_score <- function(data) {
  game <- data %>%
    group_by(player) %>%
    summarize(score = sum(points), .groups = "drop")
  
  # Handle the "shoot the moon" scenario 
  # where one player has all the points
  if (any(game$score == 26)) {
    game <- game %>%
      mutate(score = 26 - score)
  }
  game
}


hearts(c("Amy", "Beth", "Cara", "Dawn")) %>%
  calc_score()
```

### Identify a probability question you can answer with your simulation

Fundamentally, my question is what the probability is that a player using relatively little strategy accidentally "shoots the moon". 

In a shoot the moon scenario, the total game points will be (n-1) times normal, where $n$ is the number of players. This function returns a logical indicator of whether or not the total number of points generated is over 26 (indicating that someone shot the moon)

```{r}
shoot_the_moon <- function(data) {
  sum(data$score) > 26
}
```

### Design a simulation to answer the question you identified

Run your simulation with an increasing number of iterations and create some sort of plot showing your results. 

Describe any issues you encountered in your simulation, and how you addressed them (or where the whole thing went off the rails, if you couldn't address everything in time). 

```{r, eval = F}
# I'm saving the results of this so that I only have to run it once...
games <- replicate(1000, hearts(c("Amy", "Beth", "Cara", "Dawn")), simplify = F) # play 1000 games of hearts

save(games, file = "games.Rdata")

games2 <- replicate(9000, hearts(c("Amy", "Beth", "Cara", "Dawn")), simplify = F) # play 1000 games of hearts

save(c(games, games2), file = "games.Rdata")
```
```{r}
load("games.Rdata")
res <- tibble(
  full_game = games, 
  score = purrr::map(full_game, calc_score),
  stm = purrr::map_lgl(score, shoot_the_moon)
)
```

From this assessment, one player shoots the moon in approximately `r sprintf("%0.2f%%", 100*mean(res$stm))` of games played using the naive strategy
